---
title: "exploratory correlation"
author: "Senna"
date: "2024-11-20"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(reshape2)
library(tidyverse)
```


```{r}
spotify_df <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-21/spotify_songs.csv')
```
## Data Cleaning

#### feel free to edit as needed, just include a note explaining the changes you made.

What Congyu did here:\

1. remove duplicated track_id. \
   There are many songs appear in this dataset more than once, under different genre group and subgenre group, to keep this analysis fair, we remove duplicated songs.
2. remove unique ID columns we are not using. \
3. In `track_album_release_date`, many of rows miss month and year, so I create a new column `track_album_release_year` only including release year.\
4. factor variables.\
5. Adjust dancibility and energy to range 0 - 100 to enlarge the effect on graph.(we can also do this on any other factors we are interested in)
6. clean data based on data dictionary.

```{r}
spotify_df <- 
  spotify_df %>% 
  distinct(track_id,.keep_all = T) %>% select(-track_id,-track_album_id,-playlist_id) %>%
  mutate(track_album_release_year = as.numeric(str_sub(track_album_release_date,1,4))) %>% 
  select(-track_album_release_date) %>% 
  mutate(playlist_name = factor(playlist_name),
         playlist_genre = factor(playlist_genre),
         playlist_subgenre = factor(playlist_subgenre)) %>% 
  mutate(danceability = danceability * 100,
         energy = energy * 100) %>% 
  mutate(key = factor(case_match(key,0 ~ "C",
                          1 ~ "C#/Db", 2 ~ "D",
                          3 ~ "D#/Eb", 4 ~ "E",
                          5 ~ "F", 6 ~ "F#/Gb",
                          7 ~ "G", 8 ~ "G#/Ab",
                          9 ~ "A", 10 ~ "A#/Bb",
                          11 ~ "B")),
         mode = factor(case_match(mode, 0 ~ "minor",1 ~ "Major")),
         speechiness = factor(case_when(speechiness < 0.33 ~ "non-speech-like music",
                                 speechiness >= 0.33 & speechiness <= 0.66 ~ "mix of speech and music",
                                 speechiness > 0.66 ~ "entirely of spoken words")))
```

```{r}
spotify_df %>% group_by(playlist_genre) %>% 
  summarise(genre_freq = n()) %>% knitr::kable()
```

```{r}
spotify_df %>% 
  mutate(playlist_genre = fct_infreq(playlist_genre,track_popularity)) %>% 
  ggplot(aes(x = playlist_genre, y = track_popularity, color = playlist_genre)) + 
  geom_boxplot()
```
We can see there is a obvious trend between song's popularity and its genre. Pop songs is the most popular genre and edm is the least.



## Question 1
```{r}
numeric_var = spotify_df[sapply(spotify_df, is.numeric)]

cor_matr = melt(cor(numeric_var, use = "complete.obs"))
```


```{r}
ggplot(cor_matr, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0, limit = c(-1, 1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Correlation Heatmap", x = "Variables", y = "Variables", fill = "Correlation")
```

```{r}
cor_pop = spotify_df |>
  select(where(is.numeric)) |> 
  select(-track_popularity) |> 
  summarise(across(everything(), ~ cor(.x, spotify_df$track_popularity, use = "complete.obs"))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "correlation") |> 
  arrange(desc(correlation))

cor_pop

```

```{r}

cor_pop|>
  ggplot(aes(x= variable, y = correlation))+
  geom_col() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "correlation with popularity", x = "Variable", y = "correlation")

```

## Question 3

### energy

```{r}
lm(track_popularity ~ energy, data = spotify_df) %>% 
  summary()
```
From this simple linear regression, we got a super small p-value, which means we can reject the null hypothesis and conclude that energy is a significant predictor for popularity.

```{r}
spotify_df %>% 
  ggplot(aes(x = energy)) + 
  geom_histogram()
```
From this plot we can see, singers prefer to produce and sing energetic (fast, loud, and noisy) music.

```{r}
spotify_df %>% group_by(playlist_genre) %>% 
  ggplot(aes(x = energy, y = track_popularity)) + 
  geom_hex()+
  facet_wrap(~ playlist_genre,nrow = 2)

```
Although, we realize that energy is a significant predictor for popularity and current singers and song producers prefer quick and loud songs. Unfortunate, we cannot see obvious relationship between songs' energy and songs' popularity.


Try k-mean

```{r}
energy_popularity <- spotify_df %>% 
  select(track_popularity,energy)

kmeans_energy <- 
  kmeans(x = energy_popularity, centers = 3)

kmeans_energy
```

```{r}
kmeans_energy_df =
  broom::augment(kmeans_energy, energy_popularity)

kmeans_energy_df %>% 
  ggplot(aes(x = energy, y = track_popularity, color = .cluster)) +
  geom_boxplot() 
```
Energy level around 80 is most likely to have higher popularity score.

```{r}
energy_popularity_genre <- spotify_df %>% 
  select(track_popularity,energy,playlist_genre)

kmeans_energy_genre_df =
  broom::augment(kmeans_energy, energy_popularity_genre)

kmeans_energy_genre_df %>% group_by(playlist_genre) %>% 
  ggplot(aes(x = energy, y = track_popularity, color = .cluster)) +
  geom_boxplot() +
  facet_wrap(~playlist_genre,nrow = 2)
```


### key

```{r}
spotify_df %>% mutate(key = fct_infreq(key)) %>% 
  ggplot(aes(x = key)) + 
  geom_bar()
```
From this plot we can see, the most commonly used key is `C#/Db`.
Because `C#/Db` keys are often used in `R&B` and `Pop` genre.
Other than that, singers love to use simpler keys without sharp key or flat key. 

```{r}
spotify_df %>% 
  ggplot(aes(x = key, y = track_popularity,fill = key,color = key)) + 
  geom_violin()
```

```{r}
spotify_df %>% group_by(playlist_genre) %>% 
  ggplot(aes(x = key, y = track_popularity, fill = key)) + 
  geom_boxplot()+
  facet_wrap(~ playlist_genre,nrow = 2)+
  theme(axis.text.x = element_text(size = 4,angle = 30))
```



### loudness

```{r}
lm(track_popularity ~ loudness, data = spotify_df) %>% 
  summary()
```
From this simple linear regression, we got a super small p-value, which means we can reject the null hypothesis and conclude that loudness is a significant predictor for popularity.

```{r}
spotify_df %>% 
  ggplot(aes(x = loudness)) + 
  geom_histogram()
```
From this plot we can see, most of the songs have loudness between -10db to 0db.

```{r}
spotify_df %>% 
  ggplot(aes(x = loudness, y = track_popularity)) + 
  geom_hex()+
  geom_smooth()
```

```{r}
spotify_df %>% group_by(playlist_genre) %>% 
  ggplot(aes(x = loudness, y = track_popularity)) + 
  geom_hex()+
  facet_wrap(~ playlist_genre,nrow = 2)

```















