---
title: "Duration"
output: github_document
date: "2024-11-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stats)
library(effectsize)
```

```{r}
spotify_df <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-21/spotify_songs.csv')
```

```{r}
#data cleaning
spotify_df <- 
  spotify_df %>% 
  distinct(track_id, .keep_all = TRUE) %>% 
  select(-track_id, -track_album_id, -playlist_id) %>% 
  mutate(
    track_album_release_year = as.numeric(str_sub(track_album_release_date, 1, 4)), 
    playlist_name = factor(playlist_name),
    playlist_genre = factor(playlist_genre),
    playlist_subgenre = factor(playlist_subgenre)
  ) %>% 
  select(-track_album_release_date) 
```
  
```{r}
#median approach
spotify_df_median <- 
  spotify_df %>%
  mutate(duration_median = case_when(  
      duration_ms >= median(duration_ms, na.rm = TRUE) ~ "Long Duration",
      duration_ms < median(duration_ms, na.rm = TRUE) ~ "Short Duration"
      )
    )
```


```{r}
#standard deviation approach
spotify_df_sd <- spotify_df %>%
  mutate(
    duration_sd = case_when(
      duration_ms < mean(duration_ms, na.rm = TRUE) - sd(duration_ms, na.rm = TRUE) ~ "Short Duration",
      duration_ms > mean(duration_ms, na.rm = TRUE) + sd(duration_ms, na.rm = TRUE) ~ "Long Duration",
      TRUE ~ "Medium Duration"
    )
  )
```

```{r}
#k-means approach
kmeans_result <- kmeans(spotify_df$duration_ms, centers = 3)

spotify_df_kmeans <- spotify_df %>%
  mutate(
    duration_kmeans = case_when(
      kmeans_result$cluster == 1 ~ "Short Duration",
      kmeans_result$cluster == 2 ~ "Medium Duration",
      kmeans_result$cluster == 3 ~ "Long Duration"
    )
  )
```

```{r}
#percentile approach
spotify_df_perc <- spotify_df %>%
  mutate(
    duration_perc = case_when(
      duration_ms < quantile(duration_ms, 0.25, na.rm = TRUE) ~ "Short Duration",
      duration_ms > quantile(duration_ms, 0.75, na.rm = TRUE) ~ "Long Duration",
      TRUE ~ "Medium Duration"
    )
  )
```


```{r}
#median approach plot
spotify_df_median %>%
  ggplot(aes(x = playlist_genre, y = track_popularity, fill = duration_median)) +
  geom_boxplot(outlier.alpha = 0.3) +
  labs(
    title = "Popularity by Genre and Duration Category (Median)",
    x = "Music Genre",
    y = "Track Popularity",
    fill = "Duration Category"
  ) 
```
We see that for some Genre songs with long duration tend to have lower popularity, for instance, edm and r&b. However, for others there is no such significant differences.
```{r}
#median approach anova test with effect size (Eta-squared)
anova_results_median <- spotify_df_median %>%
  group_by(playlist_genre) %>%
  summarise(
    anova = list(aov(track_popularity ~ duration_median, data = cur_data())),
    eta_squared = map_dbl(anova, ~ eta_squared(.)$Eta2[1])  
  )

anova_summary_median <- anova_results_median %>%
  mutate(
    f_statistic = map_dbl(anova, ~ summary(.)[[1]]$`F value`[1]),
    p_value = map_dbl(anova, ~ summary(.)[[1]]$`Pr(>F)`[1])
  )

anova_summary_median
```
By running anova test and compute effect size using Eta-squared we see that while p-value is significant for all genre except rock with 0.45, the effect size are all small. Hence we can say that Long-duration songs are slightly less popular in some genres (e.g., EDM, R&B, Rap), but the effect is small and not meaningful for practical applications.

```{r}
#standard deviation approach plot
spotify_df_sd %>%
  ggplot(aes(x = playlist_genre, y = track_popularity, fill = duration_sd)) +
  geom_boxplot(outlier.alpha = 0.3) +
  labs(
    title = "Popularity by Genre and Duration Category (Standard Deviation)",
    x = "Music Genre",
    y = "Track Popularity",
    fill = "Duration Category"
  ) 
```
We see that for some Genre songs with long duration tend to have lower popularity, for instance, edm， pop, rap, and r&b. However, for latin and rock there is no such significant differences.

```{r}
#standard deviation approach anova test
anova_results_sd <- spotify_df_sd %>%
  group_by(playlist_genre) %>%
  summarise(
    anova = list(aov(track_popularity ~ duration_sd, data = cur_data())),
    eta_squared = map_dbl(anova, ~ eta_squared(.)$Eta2[1]))

anova_summary_sd <- anova_results_sd %>%
  mutate(
    f_statistic = map_dbl(anova, ~ summary(.)[[1]]$`F value`[1]),
    p_value = map_dbl(anova, ~ summary(.)[[1]]$`Pr(>F)`[1])
  )
anova_summary_sd
```
While all the genre music are statistically significant differences, the effect sizes are very small or even negligible (with Eta-squared < 0.02), indicating that duration explain only a small proportion of the variance in popularity. 

```{r}
#k-means approach plot
spotify_df_kmeans %>%
  ggplot(aes(x = playlist_genre, y = track_popularity, fill = duration_kmeans)) +
  geom_boxplot(outlier.alpha = 0.3) +
  labs(
    title = "Popularity by Genre and Duration Category (K-Means)",
    x = "Music Genre",
    y = "Track Popularity",
    fill = "Duration Category"
  ) 
```
We see that for some Genre songs with long duration tend to have lower popularity, for instance, edm, latin, pop, and r&b. However, for rap and rock there is no such significant differences.
```{r}
#k-means approach anova test
anova_results_kmeans <- spotify_df_kmeans %>%
  group_by(playlist_genre) %>%
  summarise(
    anova = list(aov(track_popularity ~ duration_kmeans, data = cur_data())),
    eta_squared = map_dbl(anova, ~ eta_squared(.)$Eta2[1]))

anova_summary_kmeans <- anova_results_kmeans %>%
  mutate(
    f_statistic = map_dbl(anova, ~ summary(.)[[1]]$`F value`[1]),
    p_value = map_dbl(anova, ~ summary(.)[[1]]$`Pr(>F)`[1])
  )
anova_summary_kmeans
```
Confirm to previous approaches, even with statistically significance difference the duration have small practical impact on song popularity across genres.

```{r}
#percentile approach plot
spotify_df_perc %>%
  ggplot(aes(x = playlist_genre, y = track_popularity, fill = duration_perc)) +
  geom_boxplot(outlier.alpha = 0.3) +
  labs(
    title = "Popularity by Genre and Duration Category (Percentiles)",
    x = "Music Genre",
    y = "Track Popularity",
    fill = "Duration Category"
  ) 
```
We see that for some Genre songs with long duration tend to have lower popularity, for instance, edm, latin, pop, and r&b. However, for rap and rock there is no such significant differences.

```{r}
#percentile approach anova test
anova_results_perc <- spotify_df_perc %>%
  group_by(playlist_genre) %>%
   summarise(
    anova = list(aov(track_popularity ~ duration_perc, data = cur_data())),
    eta_squared = map_dbl(anova, ~ eta_squared(.)$Eta2[1]))

anova_summary_perc <- anova_results_perc %>%
  mutate(
    f_statistic = map_dbl(anova, ~ summary(.)[[1]]$`F value`[1]),
    p_value = map_dbl(anova, ~ summary(.)[[1]]$`Pr(>F)`[1])
  )
anova_summary_perc
```
Again, while the popularity across duration categories are statistically significant difference, duration explain a small proportion of the variance in popularity.

### Overall
Long duration songs are slightly less likely to be more popular, but the effect of duration is small and varies significantly across genres.